---
title: "p8105_hw2_zx2482"
author: "Allison Xia"
date: "2023-10-02"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
options(scipen=999)
```

### Problem 1

#### Clean `pols`
```{r clean_538_pols}
month_df = 
  tibble(
    month_num = 1:12,
    month_abb = month.abb,
    month = month.name
  )

pols = 
  read_csv("./data/fivethirtyeight_datasets/pols-month.csv") |>
  separate(mon, into = c("year", "month_num", "day"), convert = TRUE) |>
  mutate(
    president = recode(prez_gop, "0" = "dem", "1" = "gop", "2" = "gop")) |>
  left_join(month_df) |>
  select(year, month, everything(), -day, -starts_with("prez")) 
```

#### Clean `snp`
```{r clean_538_snp}
snp = 
  read_csv("./data/fivethirtyeight_datasets/snp.csv") |>
  separate(date, into = c("month", "day", "year"), convert = TRUE) |>
  arrange(year, month) |>
  mutate(month = month.name[month]) |>
  select(year, month, close) 
```
#### Clean `unemployment` 
```{r clean_538_unemp}
unemployment = 
  read_csv("./data/fivethirtyeight_datasets/unemployment.csv") |>
  rename(year = Year) |>
  pivot_longer(
    Jan:Dec, 
    names_to = "month_abb",
    values_to = "unemployment"
  ) |> 
  left_join(month_df) |> 
  select(year, month, unemployment)
```

Now we merge the three datasets!

```{r merge_538}
data_538 = 
  left_join(pols, snp) |>
  left_join(unemployment)

str(data_538)
```

The `pols` data has `r nrow(pols)` observations and `r ncol(pols)` variables and tells us about the party affiliation distribution (democrat or republican) for governors and senators for a given year from years `r range(pols$year)[1]` to `r range(pols$year)[2]`. It also tells us whether the sitting president was a democrat or republican. The `snp` data has `r nrow(snp)` observations and `r ncol(snp)` variables, ranging from years `r range(snp$year)[1]` to `r range(snp$year)[2]`. The `unemployment` data has `r nrow(unemployment)` observations and `r ncol(unemployment)` variables ranging from years `r range(unemployment$year)[1]` to `r range(unemployment$year)[2]`. In Januarys in or after 1975 in which a democrat was president, the **average unemployment rate was `r filter(data_538, month == "January", year >= 1975, president == "dem") |> pull(unemployment) |> mean() |> round(2)`**.  The average unemployment rate over the same time period in which a republican was president was `r filter(data_538, month == "January", year >= 1975, president == "gop") |> pull(unemployment) |> mean() |> round(2)`.

## Problem 2
```{r clean Mr._Trash_Wheel}
mtw = 
  read_excel("./202207 Trash Wheel Collection Data.xlsx", sheet="Mr. Trash Wheel", range =cell_cols("A:N")) |> 
  janitor::clean_names() |> 
  separate(date, into = c("year_2", "month_num", "day"), convert = TRUE) |> 
  select(-year_2) |> 
  mutate(homes_powered = weight_tons * 500/30,
         year = as.numeric(year)) |> 
  mutate(wheel = "mtw") |> 
  filter(dumpster != "NA")
```

```{r clean Professor_Trash_Weel}
ptw = 
  read_excel("./202207 Trash Wheel Collection Data.xlsx", sheet="Professor Trash Wheel", range =cell_cols("A:M")) |> 
  janitor::clean_names() |> 
  separate(date, into = c("year_2", "month_num", "day"), convert = TRUE) |> 
  select(-year_2) |> 
  mutate(homes_powered = weight_tons * 500/30,
         year = as.numeric(year)) |> 
  mutate(wheel = "ptw") |> 
  filter(dumpster != "NA")
```

```{r clean Gwynnda_Trash_Wheel}
gtw = 
  read_excel("./202207 Trash Wheel Collection Data.xlsx", sheet="Gwynnda Trash Wheel", range =cell_cols("A:K")) |> 
  janitor::clean_names() |> 
  separate(date, into = c("year_2", "month_num", "day"), convert = TRUE) |> 
  select(-year_2) |> 
  mutate(homes_powered = weight_tons * 500/30,
         year = as.numeric(year)) |> 
  mutate(wheel = "gtw") |> 
  filter(dumpster != "NA")
```

```{r}
data_trash = bind_rows(gtw,mtw,ptw)
str(data_trash)
```
The `mtw` data has `r nrow(gtw)` observations and `r ncol(gtw)` variables, the he `ptw` data has `r nrow(ptw)` observations and `r ncol(ptw)` and tells us about litters the Mr. Trash Wheel, Professor Trash Wheel, and the Gwynnda Trash Wheel has removed as a water wheel vessel from the Inner Harbor in Baltimore, Maryland from years `r range(gtw$year,mtw$year,ptw$year)[1]` to `r range(gtw$year,mtw$year,ptw$year)[2]`. The total weight of trash collected by Professor Trash Wheel is `r filter(data_trash, wheel == "ptw") |> pull(weight_tons) |> sum() |> round(2)` tons. In July of 2021, the total number of cigarette butts collected by Gwynnda is `r filter(data_trash, wheel == "gtw") |> pull(cigarette_butts) |> sum () |> round(2)` 